Measuring contrast processing in the visual system using SSVEP

### 2024-05-06 review.qmd

# Conclusions

>>>>

# Future directions
The SSVEP is a powerful tool for studying contrast processing. It provides a high SNR readout of neuronal activity that is unambigously linked to the input. It is sensitive to both the amplitude and phase of the input and when combined with source imaging, it can be extracted from different cortical regions allowing researchers to track contrast processing computations across the visual pathway. Because of the high SNR, it can be measured in subjects where long recording durations are impractical (for example, infants or patients with neurological disorders) and at frequencies high enough to be effectively invisible [REF - Some high frequency work... - Romanian group]. The SSVEP is also able to 'fingerprint' modulators of the inputs through the harmonic and intermodulation terms they generate in the output. In principle, each nonlinearity in the visual pathway can be identified by its contribution to the frequency spectrum at different recording locations. This, it turn, allows researchers to study how contrast processing depends on spatial and temporal context and as well as changes in task, cognitive and behavioural state and arousal.

Although visual neuroscience is a relatively old field, there are still outstanding questions relating to contrast processing that could be addressed by SSVEP methods. For example, it is still not completely clear how contrast signals are computed in the human retina: although we have more than a century of electrophysiological data from animals, and the broad structure of cone inputs to retinal ganglion cells is understood [@LiFieldNeuron2014] we are still discovering new aspects of retinal processing that could influence the 'coining' of the visual system's currency [@gollischMeister2010, @WangPuthussery2023Nature, @uprety2022melanopsin]. The ability to frequency tag both inputs (for example, cone-directed luminance contrast) and modulators (for example, stimuli that selectively drive the intrinsically-photosensitive retinal ganglion cells) is a powerful tool to explore this first stage of image generation.

Later in the visual pathway, we would like to know more about the role of corticothalamic feedback in the LGN. Although it is commonly thought of as a simple relay between the eye and the cortex,  the majority of inputs to the LGN come from cortex not the eye, and there is good evidence that contrast processing in the LGN can be altered by top-down signals including attention [@BriggsUsrey2007, @ShermanGuillery1996, @OconnorNatNeuro2002, @GouwsJNS2014]. Bottom-up inputs to the LGN are segregated by eye and so responses there are often considered to be purely monocular but it is possible that feedback signals allow some binocular computations such as interocular normalisation [@ManyBakerRefs and others] to begin even at this relatively early stage. Frequency tagging inputs to different eyes or in different precortical pathways allows us to address neurons in different parts of the LGN. These techniques combined with advances in recording technology (for example, sensitive source-imaged EEG and MEG recordings that can resolve subcortical structures [@Attal2012, @AttalPlos12013, @Tesche1996], noninvasive deep-brain stimulation techniques [@MohammadjavadietalSciRep2022] or implanted electrode arrays [@KrolakSalmon2003]) may allow us to study these computations in more detail.

Finally...

(Adaptation, Amblyopia, Neurological disease, Chromatic channels and combinations, joint tuning for color/spatiotemporal frequncy/eye), contributions to gain pools (e.g. Solomon and Lennie 2006), inheritance of freqeuncy tags in higher areas, combination of low-level contrast features to make complex stimuli...)

*2: Sweeps and CRFs. Measurements of contrast sensitivity, extrapolating
the sweep to zero response to get t'hold. Infants and adults?*

*3: Measurements of modulation. Figures adapted from other papers:
Attention to space, attention to features, adaptation (?).
masking/surround suppression\\*

*4: Clinical (Porciatti / Tsai/ Marmite/ Amblyopia / PD?)*

*5: Future directions (decoding in frequency domain?, animals? BCI?)*

[]{#anchor-2}History

[]{#anchor-3}1. The basics of SSVEP and contrast sensitivity including a
history of both fields

SSVEP (Steady-State Visual Evoked Potential): A continuous electrical
response evoked in the brain by visual stimuli flickering at a constant
frequency (Regan, 1966).

Contrast Sensitivity: The ability to detect differences in luminance
between an object and its background (Campbell \& Green, 1965).

Regan, D. (1966). Some characteristics of average steady-state and
transient responses evoked by modulated light. Electroencephalography
and clinical neurophysiology, 20(3), 238-248.

Campbell, F. W., \& Green, D. G. (1965). Optical and retinal factors
affecting visual resolution. J Physiology, 181, 576-593.

Check Regan paper for earlier (e.g. EEG refs). Norcia review will be
helpful!

(From Tyler / Levi / Apkarian paper):

6\. Regan, D.: Rapid methods for refracting the eye and assessing the
visual acuity in amblyopia using steady-state visual evoked potentials.
In Desmedt, J.E., editor: Visual Evoked Potentials in Man: New
Developments, Oxford, 1977, Clarendon Press, pp. 418-426.

7\. Fricker, S.J.: Narrow-band filter techniques for the detection and
measurement of evoked responses, Electroencephalogr. Clin. Neurophysiol.
14:411, 1962.

8\. Van der Tweel, L. H., Sem-Jacobsen, C.W., Kamp, A., Van Leeuwen,
W.S., and Verings, F.T.H.: Objective determination of response to
modulated light, Acta Physiol. Pharmacol. Neerl. 7:528, 1958.

9\. Regan, D.: Latencies of evoked potentials to flicker and to pattern
speedily estimated by simultaneous stimulation method,
Electroencephalogr. Clin. Neurophysiol. 40:654, 1976.

10\. Tyler, C.W., Apkarian, P., and Nakayama, K.: Multiple spatial
frequency tuning of electrical responses from the human\....

\[Around here perhaps a section about what gain control is, mentioning
other methods as well including psychophysics, MRI, electrophysiology,
and other EEG markers including ERPs and evoked gamma band oscillations.
Maybe outline the Heeger gain control model and its cousins.\]

\[Sure. But in fact gain control is only a small part of this story -
especially in the early days. They were, I think, more interested in
using SSVEP to measure absolute contrast sensitivity and to get lower
bounds on things like infant visual development. Gain control might be
better as a separate section later.\]

[]{#anchor-4}2. Spatial, temporal frequency and contrast sensitivity
measurements

Basically using SSVEP to measure a cortical output amplitude for any
given input contrast. You can vary parameters like SF, TF, position,
color and of course contrast. Early on people realised that you can
'sweep' the stimulus to get a CRF. You broadly get a line in log
contrast space if you do that (Tyler) - then you can extrapolate that
line down to zero response to estimate the threshold. That doesn't
\>quite\< work but it's pretty close.

You can also use this for 'difficult' populations like babies. One
interesting story was about how SSVEP become a replacement for
preferential looking (which was the other way of looking at infant
visual development). See e.g. Davida Teller. SSVEPs allowed people to
make objective measurements of contrast sensitivity development and
deduce that the visual system was more mature (e.g. more functional) in
infancy than previously expected. Also measures of colour sensitivity.
Tinyeyes is based off those measurements. Other people: Tyler, Norcia,
Gunilla H-P, many of the people at SKERI in the 1980s and 90s. Norcia
86,88,90 - mentioned in Regan's nice autobiography :

In parallel of course, people were using frequency tagging to do single
unit work - the 1F vs 2F simple/complex cell classification scheme was
all about this (Lennie and others).

[]{#anchor-5}3. SSVEP in functional localization

Techniques like fMRI have been combined with SSVEP to achieve more
precise spatial localization (Di Russo et al., 2007). See however the
Ales cruciform paper.

Ales and Norcia:\... (showed that people's intuition about V1 upper /
lower v.f reversing polarity wasn't really correct). This feels a little
outside our scope though\... We have also combined SSVEP with source
imaging techniques to probe responses in different cortical locations
(again, Ales papers, some early stuff from Stan K? Appelbaum, Wade,
Norcia figure/ground\...) and then a host of later work from that lab
and others.

[]{#anchor-6}4. SSVEP and contrast gain control including adaptation,
masking, and attention

\[The story goes: Up to the lte 90s people were primarily interested in
measuring contrast sensitivity - the shape of the response function was
assumed to be basically log-linear - and they fit it with straight lines
to extrapolate back to zero response. But then (once Heeger's 1992 paper
had sunk in - see also stuff like Shapley and Victor 1981 [(Shapley \&
Victor, 1981)](https://www.zotero.org/google-docs/?zAmtgj)), people
started thinking about gain control - Candy and Norcia in about 1999,
Porciatti, probably a load of Tyler papers that I don't even know
about\...\]. Probably look in p

And then people worked out that if you can use SSVEP to measure contrast
responses, you can also use it to measure things that modulate contrast
responses. These include adaptation, masking, suppression, attention
(feature and space), clinical things.

[]{#anchor-7}Adaptation:

Continuous exposure to high-contrast patterns reduces contrast
sensitivity, which can be measured using SSVEP (Ross et al., 1989).
Others? Baker recent gain control paper is perhaps worth mentioning here
as a 'confound' of sorts. Engel 2018 [*(Vergeer et al.,
2018)*](https://www.zotero.org/google-docs/?kdURiS). This paper is
interesting [*(Rideaux et al.,
2023)*](https://www.zotero.org/google-docs/?J2EO3d) and \>sort\< of
SSVEP.

[]{#anchor-8}Masking:

High contrast masks can suppress the visibility of low contrast
patterns, which has implications in SSVEP amplitude (Haynes et al.,
2003).

Attention: Directing attention can enhance contrast sensitivity, as
shown in studies using SSVEP (Müller et al., 2006). Also Tsai
(dynamics), Baker / Wade (several), Winawer? I think JW has a nice
dynamic model of normalization with some MEG data. Busse et al cat/human
comparison. Candy and Norcia 2001 JNS [(Candy et al.,
2001)](https://www.zotero.org/google-docs/?7hhs5Z)

Ross, J., Speed, H. D., \& Morgan, M. J. (1989). The effects of
adaptation and masking on incremental thresholds for contrast. Vision
research, 29(2), 205-215.

Haynes, J. D., Roth, G., Stadler, M., \& Heinze, H. J. (2003).
Neuromagnetic correlates of perceived contrast in primary visual cortex.
Journal of Neurophysiology, 89(6), 2655-2666.

Müller, M. M., Picton, T. W., Valdes-Sosa, P., Riera, J.,
Teder-Sälejärvi, W. A., \& Hillyard, S. A. (2006). Effects of spatial
selective attention on the steady-state visual evoked potential in the
20--28 Hz range. Cognitive Brain Research, 24(1), 1-13.

5\. Clinical implications

[]{#anchor-9}Clinical applications (this is a whole section)

Not sure exactly how SSVEP used in clinic. mfVEP?

, e.g., in monitoring visual impairments, tracking neuronal diseases, or
neurofeedback (Norcia et al., 2015).

Tsai epilepsy [(Tsai et al.,
2011)](https://www.zotero.org/google-docs/?AdhH3f). Other photosensitive
epilepsy: Porciatti [(Porciatti et al.,
2000)](https://www.zotero.org/google-docs/?UTBQbV)

Citation:

Migraine (Regan et al). Autism?

Norcia, A. M., Appelbaum, L. G., Ales, J. M., Cottereau, B. R., \&
Rossion, B. (2015). The steady-state visual evoked potential in vision
research: A review. Journal of vision, 15(6), 4-4.

Animal work: Flies (Eliott, West, Himmelberg, Ales, Norcia): Mice/rats
(probably many - can't think off the top of my head - we were working
with a mouse EEG person at UCSF about 12 years ago\...), Monkeys
(Kiorpes?),

[]{#anchor-10}6. Future directions

Use as a readout of modulations. TMS? FUS? Marmite B12 / Fluoxetine /
amblyopia in Rats,

GABA Huang

There is a lot of SSVEP interest these days because of BCIs. I think
it's pretty weak but there is \>so much\< of it that it might be worth
mentioning\...

Advanced signal processing techniques and machine learning can be
integrated to improve SSVEP-based systems (Zhu et al., 2010).

Exploring new clinical and diagnostic applications, understanding
neurological diseases, and developing novel therapeutic interventions.

Citation:

Zhu, D., Bieger, J., Molina, G. G., \& Aarts, R. M. (2010). A survey of
stimulation methods used in SSVEP-based BCIs. Computational intelligence
and neuroscience, 2010.
